A Method for Regularization of Evolutionary Polynomial Regressions

### 2013-12-21 texts/submission_02-asc/paper_spfga-review.tex
Completed text revision in section 3.

%
%	DOCUMENT COMMANDS
%
\newcommand{\sample}[1]{\ensuremath{^{\left(#1\right)}}}
\newcommand{\at}[1]{\ensuremath{\!\left(#1\right)}}
\newcommand{\note}[1]{\textbf{\small [Note\footnote{\textbf{Note:} #1}]}}
\newcommand{\nth}{\ensuremath{^{th}}}
%

@@

Polynomials, one of the most studied subjects in mathematics, generalize li\-ne\-ar functions and define, perhaps, the simplest and most used nonlinear models. Applications include colorimetric calibration \citep{Mendes:2005aa}, explicit formul\ae\ for turbulent pipe flows
 \citep{Davidson:1999aa}, computational linguistics \citep{Sanchez:2009aa} and more recently analytical techniques for cultural heritage materials \citep{Csefalvayova:2010aa}, liquid epoxy moulding process \citep{Chan:2011aa}, B-spline surface reconstruction \citep{Galvez:2012aa}, product design \citep{Chan:2012aa} or forecasting cyanotoxins presence in water reservoirs \citep{Garcia-Nieto:2013aa}. These examples not only illustrate the wide spectrum of applications but, additionally, in each one uses, at some point, a \ac{GA}.

\acp{GA} were, arguably, one of the hottest topics of research in the recent decades and with good reason since they outline an optimization scheme easy to conceptualize and with very broad application. If a nonlinear (or otherwise) model requires parameterization, \acp{GA} provide a simple and often effective approach to search for locally optimal parameters. Research related to genetic algorithms abound and spans from the 1950s seminal work of Nils Aall Barricelli \citep{Barricelli:1962aa} in the Institute for Advanced Study of Princeton to today's principal area of study for thousands of researchers, covered in hundreds of conferences, workshops and other meetings. Perhaps the key impulse to \acp{GA} come from John Holland's work and his book ``Adaptation in Natural and Artificial Systems'' \citep{Holland:1975aa}.

@@

This paper describes an extension of the general \ac{EPR} method to find a regularized polynomial regression of a given dataset. The optimal regression results from a cost function that accounts for both the \ac{rmse} and a regularization factor to avoid overfitting. \note{Explain a bit more the overfitting avoiding method.}

@@

The second problem is treated in the \ac{GA} setting: Let $\mathcal{D}$ be a dataset as above and $\mathcal{P}$ a set of polynomials. For each polynomial $p\in \mathcal{P}$ let $\mathcal{Q}_p$ be the set of monomials in $p$ (without the coefficients) and define the (anti) fitness
%
\begin{eqnarray*}
\phi_p &=& \min_\Theta J_{\textrm{fit}}\left(\Theta;\mathcal{Q}_p,\mathcal{D}\right)
\end{eqnarray*}
%
by solving the first problem. With a fitness of every instance, the \ac{GA} genetic operators (usually mutation and crossover) evolve the population $\mathcal{P}$ until a reasonable approximation of a local minimum is found. 
%
\begin{algorithm}[t]
\begin{algorithmic}
\Function{EPR}{$D,pop_0,\epsilon, maxiter$}
	\State $pop \gets pop_0;\: err \gets 1.0+\epsilon$
	\While{$err > \epsilon \land iterations < maxiter$}
		\State $pop \gets \Call{IterateGA}{pop}$
		\State $pop \gets \Call{Sort}{pop, key = J}$\Comment{Sort population by regression error}
		\State $err \gets J\left(\Call{First}{pop}\right)$
	\EndWhile
	\State\Return $\Call{First}{pop}$
\EndFunction
\end{algorithmic}
\caption{This \ac{EPR} algorithm uses linear regression for the calculation of the \ac{rmse} $J$ and the space of polynomials is searched in the \acp{GA} iteration step. At exit the \ac{rmse} of the fittest instance is bounded by $\epsilon$ or the maximum number of allowed iterations.}\label{alg:gapoly}
\end{algorithm} 
%
Notice that the properties of \acp{GA} and linear regression entail that Algorithm \ref{alg:gapoly} converges to a polynomial that is a local minimum of the fitness function, encapsulated in the \ac{rmse} function $J_{\textrm{fit}}$.
%

@@

The polynomial regression \ac{rmse} considered so far accounts for the ability to predict the transformed test-set. A known problem of using a cost function based only in the dataset \ac{rmse} (and of polynomial regressions in general) is the tendency to overfit  training data.

Excessive variance of the estimation method can be reduced by regularizing the \ac{rmse} function with a penalty factor. Thus, to reduce polynomial complexity and variance by regularizing the size of the monomial set the \ac{rmse} function from equation \ref{eq:rmse} is multiplied by a factor $\lambda^{k}$
\begin{eqnarray}
J_{\textrm{reg}}\left(\Theta, \lambda;\mathcal{Q},\mathcal{D}\right) &=& \lambda^{k} J_{\textrm{fit}}\left(\Theta;\mathcal{Q},\mathcal{D}\right)\label{eq:rmse-reg}
\end{eqnarray}
%
where $k$ is the number of monomials in the polynomial. When $\lambda > 1$ polynomials with more monomials are penalized. The regularized extension of \ac{EPR} is denoted by \acf{EPRR}.

@@

The results were found using \texttt{R} programming language~\citep{R-Core-Team:2013aa}\footnote{The datasets and the \texttt{R} code used to produce the results and plots in this paper are available online at \url{https://github.com/jpneto/GenAlgPoly}.}. To compare this paper's proposed algorithm we applied the exact same train and test samples using several well-known learning algorithms for regression, namely: the classic \ac{EPR}, Linear Regression, Support Vector Machines~\citep{Meyer:2012aa}, Regression Trees~\citep{Therneau:2013aa} and Conditional Inference Trees~\citep{Hothorn:2006aa,Strobl:2007aa,Strobl:2008aa}. To achieve better error results, the \ac{SVM} and Regression Tree algorithms had their parameters tuned.

In order to train and test the performance of \ac{EPRR} several mainstream datasets were used. For each dataset, we selected 70\% for training purposes and the remaining observations to make a test set in order to compute the estimated \ac{rmse}. To achieve more robust results, each dataset were processed $25$~times, each one with different samples for the train and test sets. For the datasets with attribute values of different magnitudes, a preliminary scaling was executed. The results below are box plots for the test set error predictions over these different runs.

\begin{figure*}[tb]\begin{center}
\includegraphics[width=0.98\textwidth]{figure_2.pdf}
\note{Plots of ``Regression Trees'' and ``Conditional Inference Trees'' look the same!}
\caption{Results for Artificial dataset. As shown, polynomial regression finds the exact polynomial structure that generates the dataset: in line 1, the error box is centered in $0$ and has width $0$.  The regression methods depicted are: 1. \ac{EPRR}, 2. Linear Regression, 3. SVM, 4. Regression Trees and 5. Conditional Inference Trees}
\label{artificial_dataset1_lambda1.0}
\end{center}\end{figure*}

\begin{description}
\item{\textsc{Artificial}} this is an artificial dataset with four numeric features, $x_1, \ldots x_4$, where $x_1,x_3$ are outcomes from Poisson random variables, and $x_2,x_4$ from Normal random variables. The dependent variable~$y$ is given by expression $x_2x_4^2 + x_1^2x_3 + 5$. The dataset includes $n=50$~observations.

This dataset was used in order to verify if \ac{EPR} was able to find the polynomial relation, which the algorithm did (cf. figure~\ref{artificial_dataset1_lambda1.0}). The genetic algorithm run with a population of $n=100$~solutions with $50$~iterations for each run.

In the next datasets, the population of the genetic algorithm had size~$n=300$ with 100~iterations for each run. \ac{EPRR} was run with regularization parameter $lambda = 0.8$.

\begin{figure*}[tb]\begin{center}
\begin{tabular}{cc}
\includegraphics[width=0.48\textwidth]{figure_3a.pdf}
%
&
%
\includegraphics[width=0.48\textwidth]{figure_3b.pdf}
%
\\
%
\includegraphics[width=0.48\textwidth]{figure_3c.pdf}
%
&
%
\includegraphics[width=0.48\textwidth]{figure_3d.pdf}
%
\end{tabular}

\caption{Summary results for different regression methods on the Housing, Abalone, Auto MPG and Kinematics. Although \ac{EPRR} not always achieves the smallest error, performance is on-par with more sophisticated methods.  The regression methods depicted in these figures are: 1. \ac{EPRR}, 2. \ac{EPR}, 3. Linear Regression, 4. SVM, 5. Regression Trees and 6. Conditional Inference Trees}
\label{Housing_dataset_lambda0.8_25runs}
\label{Abalone_dataset_lambda0.8_25runs}
\label{Auto-Mpg_dataset_lambda0.8_25runs}
\label{Kinematics300_lambda0.8_25runs}
\end{center}\end{figure*}

\item{\textsc{Housing}}: This data set concerns the task of predicting housing values in areas of Boston. There are $m=13$ continuous attributes and the dependent variable is the median value of owner-occupied homes in \$1000's. There are $n=506$ observations.

\item{\textsc{Abalone}} This dataset can be used to predict the age of a abalone shell using the given $m=8$ numeric attributes concerning several physical measurements. There are $n=4177$ observations.

@@

\item{\textsc{Auto MPG}} This dataset is used to predict fuel consumption in miles per gallon, based on two discrete and five continuous attributes ($m=7$). There are $n=398$ observations.

@@

\item{\textsc{Kinematics}} This dataset is concerned with the realistic simulation of the forward kinematics of an 8 link robot arm. The task is to predict the distance of the end-effector from a target using $m=8$ continuous attributes. There are $n=8192$ observations.

@@

\end{description}

\subsection{Convergence speed}

The \ac{GA} quickly proceeds in the first~50 to~100 generations to reasonable error rates. Then, it proceeds slower achieving best solutions with marginal error reduction. Since the entire learning process takes some time, in the current R implementation, placing a limit between 50 to 100 generations already achieves good results, relative to higher iteration values. Figure~\ref{Abalone_fitnessProgress} shows a typical error evolution for the dataset Abalone given a regularized version ($\lambda = 0.975$) and a non-regularized ($\lambda = 1.0$).

\begin{figure*}[tb]
\begin{center}
\includegraphics[width=0.98\textwidth]{figure_4.pdf}
\caption{Error progress for Abalone dataset during a single execution of the genetic algorithm. The figure shows the fitness evolution for two different regularization values. The population for both consisted of $200$ polynomials. The error values seem to stabilize around iteration~250.}
\label{Abalone_fitnessProgress}

\end{center}
\end{figure*}

@@

Of the regression methods considered, \ac{SVM} achieved the best results in 3 out of 4 datasets. However, \ac{SVM} was tuned for each particular dataset while \ac{EPRR} was executed with its parameters preset. Even so, \ac{EPRR} had the best error in the Abalone dataset, and achieved competitive results in the Housing and Auto MPG datasets.

Comparing \ac{EPR} and \ac{EPRR} -- which is the main article's topic -- the regularized version achieved much better results at Abalone and especially Kinematics. The Housing improved errors when compared with \ac{EPR} in a difference in means, resulted in a 95\% HDI (Highest Density Interval) equal to [0.001, 0.119] which, while borderline, achieves statistical significance. Only in the Auto MPG dataset \ac{EPR} achieved better results, even if not that different from \ac{EPRR}.

For complexity considerations \ac{EPR} and \ac{EPRR} demands some processing time. On a quad-core computer, processing the Kinematics dataset (with 8k~observations) takes approximately 5~minutes. The processing time can probably be speeded between one to two orders in magnitude if the process is implemented in a low level programming language like \texttt{C++}. However, speed optimization was not the focus of this article.

>>>>

%
%	DOCUMENT COMMANDS
%
\newcommand{\sample}[1]{\ensuremath{^{\left(#1\right)}}}
\newcommand{\at}[1]{\ensuremath{\!\left(#1\right)}}
\newcommand{\note}[1]{\textbf{\small{[Note: {#1}]}}}
\newcommand{\nth}{\ensuremath{^{th}}}
%

@@

Polynomials, one of the most studied subjects in mathematics, generalize li\-ne\-ar functions and define, perhaps, the simplest and most used nonlinear models. Applications include colorimetric calibration \citep{Mendes:2005aa}, explicit formul\ae\ for turbulent pipe flows
 \citep{Davidson:1999aa}, computational linguistics \citep{Sanchez:2009aa} and more recently analytical techniques for cultural heritage materials \citep{Csefalvayova:2010aa}, liquid epoxy moulding process \citep{Chan:2011aa}, B-spline surface reconstruction \citep{Galvez:2012aa}, product design \citep{Chan:2012aa} or forecasting cyanotoxins presence in water reservoirs \citep{Garcia-Nieto:2013aa}. These examples not only illustrate the wide spectrum of applications but, additionally, each one uses, at some point, \ac{GA}.

Evolutionary algorithms, including \ac{GA}, were, arguably, one of the hottest topics of research in the recent decades and with good reason since they outline an optimization scheme easy to conceptualize and with very broad application. If a nonlinear (or otherwise) model requires parameterization, \acp{GA} provide a simple and often effective approach to search for locally optimal parameters. Related research abound and spans from the 1950s seminal work of Nils Aall Barricelli \citep{Barricelli:1962aa} in the Institute for Advanced Study of Princeton to today's principal area of study for thousands of researchers, covered in hundreds of conferences, workshops and other meetings. Perhaps the key impulse to \acp{GA} come from John Holland's work and his book ``Adaptation in Natural and Artificial Systems'' \citep{Holland:1975aa}.

@@

This paper describes an extension of the general \ac{EPR} method to find a regularized polynomial regression of a given dataset. The optimal regression results from a cost function that accounts for both the \ac{rmse} and a regularization factor to avoid overfit by penalysing polynomial complexity.

@@

The second problem is treated in the \ac{GA} setting: Let $\mathcal{D}$ be a dataset as above and $\mathcal{P}$ a set of polynomials. For each polynomial $p\in \mathcal{P}$ let $\mathcal{Q}_p$ be the set of monomials in $p$ (without the coefficients) and define the (anti) fitness
%
\begin{eqnarray*}
\phi_p &=& \min_\Theta J_{\textrm{fit}}\left(\Theta;\mathcal{Q}_p,\mathcal{D}\right)
\end{eqnarray*}
%
by solving the first problem. With a fitness of every instance, the \ac{GA} genetic operators (usually mutation and crossover) evolve the population $\mathcal{P}$ until a reasonable approximation of a local minimum is found. 
%
\begin{algorithm}[t]
\begin{algorithmic}
\Function{EPR}{$D,pop_0,\epsilon, maxiter$}
	\State $pop \gets pop_0;\: err \gets 1.0+\epsilon$
	\While{$err > \epsilon \land iterations < maxiter$}
		\State $pop \gets \Call{IterateGA}{pop}$
		\State $pop \gets \Call{Sort}{pop, key = J}$\Comment{Sort population by regression error}
		\State $err \gets J\left(\Call{First}{pop}\right)$
	\EndWhile
	\State\Return $\Call{First}{pop}$
\EndFunction
\end{algorithmic}
\caption{This \ac{EPR} algorithm uses linear regression for the calculation of the \ac{rmse} $J$ and the space of polynomials is searched in the \acp{GA} iteration step. At exit the \ac{rmse} of the fittest instance is bounded by $\epsilon$ or the maximum number of allowed iterations.}\label{alg:gapoly}
\end{algorithm} 
%
The properties of \acp{GA} and linear regression entail that Algorithm \ref{alg:gapoly} converges to a polynomial that is a local minimum of the fitness function, encapsulated in the \ac{rmse} function $J_{\textrm{fit}}$.
%

@@

The polynomial regression \ac{rmse} considered so far accounts for the ability to predict the transformed testset. A known problem of using a cost function based only in the dataset \ac{rmse} (and of polynomial regressions in general) is the tendency to overfit  training data. Excessive variance of the estimation method can be reduced by regularizing the \ac{rmse} function with a penalty factor. Thus, to reduce polynomial complexity and variance by regularizing the size of the monomial set the \ac{rmse} function from equation \ref{eq:rmse} is multiplied by a factor $\lambda^{k}$
\begin{eqnarray}
J_{\textrm{reg}}\left(\Theta, \lambda;\mathcal{Q},\mathcal{D}\right) &=& \lambda^{k} J_{\textrm{fit}}\left(\Theta;\mathcal{Q},\mathcal{D}\right)\label{eq:rmse-reg}
\end{eqnarray}
%
where $k$ is the number of monomials in the polynomial. When $\lambda > 1$ polynomials with more monomials are penalized. The regularized extension of \ac{EPR} is denoted by \acf{EPRR}.

@@

Here is described the experiment setup used to gather and summarize the empirical evidence that supports this comparative study of \ac{EPR} and \ac{EPRR}. Evaluation is focused in \ac{rmse} distribution and, besides \ac{EPR} and \ac{EPRR}, also uses several common regression methods and datasets easily accessible in \texttt{R}, the free software environment for statistical computing and graphics~\citep{R-Core-Team:2013aa}\footnote{The datasets and \texttt{R} code used to produce the results and plots in this paper are available online at \url{https://github.com/jpneto/GenAlgPoly}.}. A small consideration on the convergence speed concludes this section.

\subsection{Regression Methods and Datasets}

The \ac{EPRR} method is ranked against several well-known learning algorithms for regression, namely: non-regularized \ac{EPR}, Linear Regression, Support Vector Machines~\citep{Meyer:2012aa}, Regression Trees~\citep{Therneau:2013aa} and Conditional Inference Trees~\citep{Hothorn:2006aa,Strobl:2007aa,Strobl:2008aa}. To achieve better error results the \ac{SVM} and Regression Tree parameters are tuned in each dataset.

The performance of each method is evaluated on several common datasets. From each dataset 70\% of the observations are reserved for training purposes and the remaining observations used to estimate the \ac{rmse}. To enhance the robustness of results this process is repeated $25$~times, each time with different shuffling of the samples in the train and test sets. Some datasets with attribute values of different magnitudes have a pre-processing scaling transformation. The box plots in figures \ref{artificial_dataset1_lambda1.0} and \ref{fig:four.datasets.summary} resume the test set error distributions over these different runs.

One of the used datasets, \textsc{Artificial}, has a special role: it is used to test if \ac{EPRR} is able to discover a polynomial model. The idea of this test is to generate a polynomial dependent variable and measure the \ac{EPRR} \ac{rmse} after fitting the dataset. The genetic algorithm parameterization for this dataset uses a population with size $n=100$ and evolves for $50$~generations. For the remaining datasets the population has size~$n=300$ and evolves for $100$~generations.

\begin{figure*}[tb]\begin{center}
\includegraphics[width=0.98\textwidth]{figure_2.pdf}
\caption{Testing polynomial discovery. The dataset is generated from a polynomial expression and, as shown, \ac{EPRR} finds the exact generator structure: in line 1, the error box is centered in $0$ and has width $0$.  The regression methods depicted are: 1. \ac{EPRR}, 2. Linear Regression, 3. SVM, 4. Regression Trees and 5. Conditional Inference Trees}
\note{Plots of ``Regression Trees'' and ``Conditional Inference Trees'' look the same!} 
\label{artificial_dataset1_lambda1.0}
\end{center}
\end{figure*}

\begin{description}
\item{\textsc{Artificial}} is a polynomial dataset with four numeric features, $x_1, \ldots x_4$, where $x_1,x_3$ are outcomes from Poisson random variables, and $x_2,x_4$ from Normal random variables. The dependent variable is given by the polynomial expression $y = x_2x_4^2 + x_1^2x_3 + 5$. The dataset includes $n=50$~observations;

\item{\textsc{Housing}} concerns the task of predicting housing values in areas of Boston. There are $n=506$ observations of $m=13$ continuous attributes and one dependent variable, the median value of owner-occupied homes in thousands of USD;

\item{\textsc{Abalone}} is used to predict the age of a abalone shell using $m=8$ numeric attributes concerning several physical measurements. There are $n=4177$ observations;

@@

\item{\textsc{Auto MPG}} gathers fuel consumption in miles per gallon, based on two discrete and five continuous attributes ($m=7$). There are $n=398$ observations;

@@

\item{\textsc{Kinematics}} results from a realistic simulation of the forward kinematics of an 8 link robot arm. The task is to predict the distance of the end-effector from a target using $m=8$ continuous attributes. There are $n=8192$ observations;

@@

\end{description}
%
\begin{figure*}[tb]\begin{center}
\begin{tabular}{cc}
\includegraphics[width=0.48\textwidth]{figure_3a.pdf}
&
\includegraphics[width=0.48\textwidth]{figure_3b.pdf}
\\
\includegraphics[width=0.48\textwidth]{figure_3c.pdf}
&
\includegraphics[width=0.48\textwidth]{figure_3d.pdf}
\end{tabular}
%
\caption{Summary results for different regression methods on diverse datasets. Although \ac{EPRR} not always achieves the smallest expected error, performance is on-par with more sophisticated methods.  The regression methods depicted in these figures are: 1. \ac{EPRR}; 2. \ac{EPR}; 3. Linear Regression; 4. \ac{SVM}; 5. Regression Trees; 6. Conditional Inference Trees;}
\label{fig:four.datasets.summary}
\label{Housing_dataset_lambda0.8_25runs}
\label{Abalone_dataset_lambda0.8_25runs}
\label{Auto-Mpg_dataset_lambda0.8_25runs}
\label{Kinematics300_lambda0.8_25runs}
\end{center}\end{figure*}

\subsection{Convergence speed}
%
\begin{figure*}[tb]
\begin{center}
\includegraphics[width=0.98\textwidth]{figure_4.pdf}
\caption{Error progress for the \textsc{Abalone} dataset during a single execution of the genetic algorithm. The figure shows the fitness evolution for two different regularization values. The population for both consists of $200$ polynomials. The error values seem to stabilize around iteration~250.}
\label{Abalone_fitnessProgress}
\end{center}
\end{figure*}
%
Since this work is oriented to the \ac{rmse} of the \ac{EPRR} model it is necessary to assess how this depends on the number of generations of the \ac{GA}. As illustrated in Figure~\ref{Abalone_fitnessProgress}, the \ac{rmse} quickly drops during the initial $50$ to $100$ generations. Then, it proceeds slower achieving better solutions only with marginal error reduction.

%Since the entire learning process takes some time, in the current \texttt{R} implementation, placing a limit between 50 to 100 generations already achieves good results, relative to higher iteration values. Figure~\ref{Abalone_fitnessProgress} shows a typical error evolution for the dataset \textsc{Abalone} given a regularized version ($\lambda = 0.975$) and a non-regularized ($\lambda = 1.0$).\note{Why not $\lambda=0.8$?}

@@

Of the regression methods considered, \ac{SVM} achieved the best results in 3 out of 4 datasets. However, \ac{SVM} was tuned for each particular dataset while \ac{EPRR} was executed with its parameters preset. Even so, \ac{EPRR} had the best error in the \textsc{Abalone} dataset, and achieved competitive results in the \textsc{Housing} and \textsc{Auto MPG} datasets.

Comparing \ac{EPR} and \ac{EPRR} -- which is the main article's topic -- the regularized version achieved much better results at \textsc{Abalone} and especially \textsc{Kinematics}. The \textsc{Housing} dataset improved errors when compared with \ac{EPR} in a difference in means, resulted in a 95\% HDI (Highest Density Interval) equal to [0.001, 0.119] which, while borderline, achieves statistical significance. Only in the \textsc{Auto MPG} dataset \ac{EPR} achieved better results, even if not that different from \ac{EPRR}.

For complexity considerations \ac{EPR} and \ac{EPRR} demands some processing time. On a quad-core computer, processing the \textsc{Kinematics} dataset (with 8k~observations) takes approximately 5~minutes. The processing time can probably be speeded between one to two orders in magnitude if the process is implemented in a low level programming language like \texttt{C++}. However, speed optimization was not the focus of this article.
